---
title: "State of the Ecosystem:"
subtitle: "New England 2019"
author: "Sean Hardison, Sarah Gaichas, Sean Lucey, Scott Large <br /> Ecosystem Dynamics and Assessment <br /> Integrated Statistics & Northeast Fisheries Science Center"
output:
  xaringan::moon_reader:
    css: ["default", "libs/20190522_Audubon_Hardison_theme.css"]
    lib_dir: libs
    nature:
      titleSlideClass: ["right", "middle", "my-title"]
      highlightStyle: githubp
      highlightLines: true
      countIncrementalSlides: false
  revealjs::revealjs_presentation:
    self_contained: false
    reveal_plugins: ["notes", "search"]
---

class: top, left

```{r setup, include=FALSE}

options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = F,
                      warning = F,
                      message = F)
#Plotting and data libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(ecodata)
library(here)
library(kableExtra)
library(ggrepel)
library(stringr)
library(patchwork)
library(grid)
library(plotly)
library(vegan)
library(rpart)
library(colorRamps)

#GIS libraries
library(sf)
library(rgdal)
library(raster)
library(rnaturalearth)

data.dir <- here::here("data")
r.dir <- here::here("R")

#Time series constants
shade.alpha <- 0.3
shade.fill <- "lightgrey"
lwd <- 1
pcex <- 2
trend.alpha <- 0.5
trend.size <- 2
hline.size <- 1
hline.alpha <- 0.35
hline.lty <- "dashed"
label.size <- 5
hjust.label <- 1.5
letter_size <- 4
feeding.guilds <- c("Apex Predator","Piscivore","Planktivore","Benthivore","Benthos")
x.shade.min <- 2009
x.shade.max <- 2018
map.lwd <- 0.4
#Function for custom ggplot facet labels
label <- function(variable,value){
  return(facet_names[value])
}

#facet names for titles
facet_names <- list("Apex predators" = expression("Apex predators"),
                    "Piscivores" = expression("Piscivores"),
                    "Planktivores" = expression("Planktivores"),
                    "Benthivores" = expression("Benthivores"),
                    "Benthos" = expression("Benthos"))
#CRS
crs <- "+proj=longlat +lat_1=35 +lat_2=45 +lat_0=40 +lon_0=-77 +x_0=0 +y_0=0 +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"

#Coastline shapefile
coast <- ne_countries(scale = 10,
                          continent = "North America",
                          returnclass = "sf") %>%
             sf::st_transform(crs = crs)

#State polygons
ne_states <- ne_states(country = "united states of america",
                                      returnclass = "sf") %>%
  sf::st_transform(crs = crs)

#high-res polygon of Maine
#new_england <- read_sf(gis.dir,"new_england")

#EPU shapefile
epu_sf <- ecodata::epu_sf %>% 
  filter(EPU %in% c("MAB","GB","GOM"))
```

## Implications for protected species: Common tern diet

.pull-left[
* Warmer waters affect vertical distribution of prey species
  * Results in more generalist diet choices
  
* High diet diversity in past decade may be the result of decreases in preferred prey type (hake, sandlance)
  * Butterfish appear in diet in warm years, are difficult to digest; may cause starvation
]
.pull-right[
```{r tern-diet-diversity, fig.height=8}
aggregate_prod <- ecodata::common_tern %>% 
    filter(!str_detect(Var, "Diet|Sum"))  %>% 
  mutate(Island = word(Var, 1),
         Var = word(Var, 3),
         Island = plyr::mapvalues(Island, from = c("EER","JI","MR","OGI","PINWR","SINWR","STI"),
                                  to = c("Eastern Egg Rock", "Jenny Island", "Matinicus Rock", "Outer Green Island", "Pond Island", "Seal Island","Stratton Island"))) %>%
  group_by(Time) %>% 
  dplyr::summarise(Mean = mean(Value, na.rm = T),
                   SE = sd(Value, na.rm = T)/sqrt(n()),
                   SD = sd(Value, na.rm = T),
                   n = n()) %>% 
  mutate(Mean = ifelse(is.na(SE),NA,Mean),
         se.low = Mean - SE,
         se.high = Mean + SE,
         hline = mean(Mean, na.rm = T))

prodplot <- aggregate_prod %>% ggplot() +
#Highlight last ten years
  annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
  geom_line(aes(x = Time, y = Mean), size = lwd-0.75) +
  geom_point(aes(x = Time, y = Mean), size = pcex-0.75) +
  geom_gls(aes(x = Time, y = Mean)) +
  geom_errorbar(aes(x = Time,
                    ymin = se.low,
                  ymax = se.high), 
                width = 0.25) +
  scale_x_continuous(expand = c(0.01, 0.01),limits = c(1991,2018)) +
  guides(color = FALSE) +
  ggtitle("Common tern productivity") +
  ylab(expression("Fledged chicks per nest")) +
  xlab("Time")+
  geom_hline(aes(yintercept = hline),
           color = "black",
           size = hline.size,
           alpha = hline.alpha,
           linetype = hline.lty) +
  labs(tag = "")  +
  theme_ts()+
  theme(axis.title = element_text(size = 14),
          plot.title = element_text(size = 16, face = "bold"))

diet_div <- ecodata::common_tern %>% 
  filter(str_detect(Var, "Diet"),
         !str_detect(Var, "Sum")) %>% 
  mutate(Island = word(Var, 1),
         Var = word(Var, 4)) %>% 
  group_by(Island, Time) %>%
  dplyr::summarise(evenness = diversity(Value)/log(specnumber(Value)),
                   shannon = diversity(Value),
                   simpson = diversity(Value, index = "simpson")) %>% 
  gather(.,Var,Value,-Island, -Time) %>% 
  group_by(Var, Time) %>%
  dplyr::summarize(Value = mean(Value, na.rm = T),
                   sd = sd(Value, na.rm = T),
                   n = n()) %>%
  group_by(Var) %>% 
  mutate(hline = mean(Value, na.rm = T))

shannon <- diet_div %>% 
  filter(Var == "shannon") %>% 
ggplot(aes(x = Time, y = Value)) +
      annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
  geom_line() +
  geom_point() +
  #geom_gls() +
  scale_x_continuous(expand = c(0.01, 0.01),limits = c(1992,2018)) +
  ggtitle("Common tern diet diversity")+
  ylab(expression("Shannon Diversity")) +
  xlab("")+
  geom_hline(aes(yintercept = hline),
           size = hline.size,
           alpha = hline.alpha,
           linetype = hline.lty) +
  theme_ts() +
  theme(axis.title = element_text(size = 14),
          plot.title = element_text(size = 16, face = "bold"))

source("R/get_tern_map.R")
tern_loc <- get_tern_map()
tern_loc + shannon + prodplot + plot_layout(ncol = 1, heights = c(1,0.4,0.4))
```
]
