---
title: "Multispecies Operating Models"
subtitle: "ICES WKMSEDEV<br /> 28 September 2023"
author: "Sarah Gaichas, Presenter<br /><br /><br />Contributors: Andy Beet, Joe Caracappa,  Gavin Fay (UMass Dartmouth), Sarah Gaichas, Robert Gamble,<br />Madeleine Guyant (UMass Dartmouth), Jerelle Jesse (GMRI/UMaine), Lisa Kerr (GMRI/UMaine),<br />Emily Liljestrand (UMass/NEFSC), Sean Lucey, Maria Cristina Perez (UMass Dartmouth)"
output:
  bookdown::word_document2: default 
---

```{r setup, include=FALSE}

options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = F,
                      fig.retina = 3,
                      warning = F,
                      message = F)
#Plotting and data libraries
library(tidyverse)
library(tidyr)
library(here)
```

```{r, load_refs, include=FALSE, cache=FALSE}
library(RefManageR)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           cite.style = "authoryear",
           longnamesfirst = FALSE,
           max.names = 1,
           style = "markdown")
myBib <- ReadBib("./mskeyrun.bib", check = FALSE)

#,
#           hyperlink = FALSE,
#           dashed = FALSE

```

# Multispecies and Ecosystem Operating Models

## Overview

Combining existing tools to evaluate EBFM procedures: Link multispecies model to MSE framework with built in single species assessment tools for New England prototype EBFM MSE (pMSE)

Fundamental objectives:
*  Maintain or increase inflation adjusted total value for the fishery
*  Preserve ecosystem function and structure
*  Maintain stock complex biomass around levels that optimize fishing opportunities 
*  Prevent overfishing
*  Response of regulations to stocks at low abundance, and recovery of depleted stocks
*  Reduce regulatory complexity

Example explanatory materials, including Rshiny for exploring results

p.s. Food web model Rpath MSE capabilities

p.p.s. Atlantis potential


# [EBFM pMSE](https://d23h0vhsm26o6d.cloudfront.net/pMSE-Final-Report2.pdf) 

  
## Gavin Fay, Lisa Kerr, Madeleine Guyant, Jerelle Jesse, Emily Liljestrand

Hydra: multispecies operating model conditioned on Georges Bank data within MSE framework as prototype test of EBFM strategies. 

Results: additional flexibility and increased yield possible with EBFM "ceilings and floors" without increased risk to single stocks.

 
# Operating model building blocks

  

## Multispecies catch at length Hydra `r Cite(myBib, "gaichas_combining_2017")`

*Hydra-Associated GitHub repositories*
* hydra-sim (Simulation Model Wiki): https://github.com/NOAA-EDAB/hydra_sim/wiki
* hydra-sim (estimation fork): https://github.com/thefaylab/hydra_sim
* hydradata (estimation fork): https://github.com/thefaylab/hydradata
* hydra-diag (diagnostics): https://github.com/thefaylab/hydra_diag

## Groundfish MSE framework `r Cite(myBib, "mazur_consequences_2023")`

*Groundfish MSE repository, branches, and tools*

* main branch (including wiki): https://github.com/lkerr/groundfish-MSE
* pMSE branch: https://github.com/lkerr/groundfish-MSE/tree/JJ-EBFM
* includes [ASAP model](https://noaa-fisheries-integrated-toolbox.github.io/ASAP) and Plan B smooth
* uses SAMtool state space surplus production assessment
 
# Setup 
  

## Management Alternatives

*  Single-species management, with stock-specific assessments and catch advice.
*  Single-species management, with stock-specific assessments and catch advice and
dynamic reference points.
*  Stock complex management, with stock-complex level assessments and abundance
index thresholds.
*  Stock complex management, with stock-complex level assessments and abundance
index thresholds, with gear-based stock complexes.
*  Stock complex management, with stock-complex level assessments and dynamic
reference points.


## Operating Model Scenarios

*  Base
*  Fleet dynamics (Adjust q for economic value)
*  Initial biomass (Below base)
*  Predator pressure (Increase M1 <30cm)
*  Prey change (Increase other food)

 

# Outreach materials

  
## Summary pages

*  pMSE Overview  
*  Management Objectives and Performance Metrics 
*  Management Alternatives
*  Results Engagement 
*  Operating Models 
*  Assessment Models


## Rshiny

https://gavinfay.shinyapps.io/results-viewer/

 

# More Potential Ecosystem Operating Models

## Atlantis  and Rpath 

 Food web responses to management?

[Rpath](https://github.com/NOAA-EDAB/Rpath) `r Cite(myBib, "lucey_conducting_2020")` with MSE capability `r Cite(myBib, "lucey_evaluating_2021")` 

[Rpath MSE course code](https://github.com/thefaylab/cinar-mse-projects/blob/main/multispecies/RpathOM_HCR_PSPS.md)


## Herring MSE food web model results: ecosystem tradeoffs

Tradeoffs between forage groups and mixed impacts to predators apparent when multiple species and full predator prey interaction feedbacks can be included

  
* Rpath Ecosense functions evaluate parameter uncertainty within a scenario

* Now we have MSE closed loop possibilities in Rpath `r Cite(myBib, "lucey_evaluating_2021")`

* Can implement HCRs with predator prey interactions 

## Herring MSE food web model results: ecosystem uncertainty?

  
Compare 10% change (green, same as previous slide gray boxes) with more extreme "herring" biomass:  
* 50% increase from base herring biomass (red)
* 50% decrease from base herring biomass (blue)

More system uncertainty with increased herring biomass?

# Atlantis NEUSv2: **major** update

  
## Joe Carracappa, Andy Beet, Robert Gamble
*  Forcing with GLORYS12V1 and primary production in the newly calibrated model `r Cite(myBib, "caracappa_northeast_2022")`
*  Running in the cloud via NOAA HPC

## In progress
*  Sensitivity to fishing scenarios
*  Testing ecosystem overfishing indicators for the Northeast US
*  Integrating spatial fleets and ports of origin
*  Climate projections using MOM-6 planned

.footnote[https://github.com/NOAA-EDAB/neus-atlantis]  

 
## References
.contrib[
```{r refs, echo=FALSE, results="asis"}
PrintBibliography(myBib)
```
]  

## Additional Resources

[EBFM pMSE June 2023 New England Council](https://www.nefmc.org/library/june-2023-ecosystem-based-fishery-management-ebfm-committee)
*  [Presentation](https://d23h0vhsm26o6d.cloudfront.net/2_Draft-final-Prototype-Management-Evaluation-pMSE-simulation-results-and-output-2nd-mailing_2023-07-06-181333_tsor.pdf)
*  [Final Report](https://d23h0vhsm26o6d.cloudfront.net/pMSE-Final-Report2.pdf)
*  [One Page Summaries](https://d23h0vhsm26o6d.cloudfront.net/4b_pMSE-one-page-sumamries_2023-07-06-181512_pgju.pdf)
*  ICES WKMSEDEV sharepoint: Background Documents



Slides available at https://noaa-edab.github.io/presentations  
Contact: <Sarah.Gaichas@noaa.gov>

